An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2011-03-16 paper.tex
Revtex 4-1 or revtex 4.
Added references, updated intro.

\documentclass[aps,prd,reprint]{revtex4-1}

@@

\bibliographystyle{apsrev4-1}

@@

Selection among alternative theoretical models given an observed data
set is an important challenge in many areas of physics and astronomy.
In a Bayesian context, model selection involves computing the evidence
for each model given the data.  The model evidence is an integral of
the unnormalized posterior probability distribution over the model
parameter space, representing the ``total probability'' in that model.
Models with larger evidence are preferred; the ratio of the evidences
of two models is the Bayes ratio between them.  The product of the
Bayes ratio and the ratio of prior probabilities for the model gives
the odds ratio for the models, the ratio of Bayesian probabilities for
the two models given the observation data.

There are many ways to compute model evidences.  In low-dimensional
parameter spaces, the unnormalized posterior probability can be
evaluated on a grid or lattice and the integral can be performed
directly.  For many problems or models of interest, however, the
dimensionality of parameter space is too large to make this approach
practical.  Nested sampling \cite{Skilling:2006,Feroz:2009} \ilya{More
  references?} is a stochastic method for exploring the parameter
space of a model while computing the evidence integral statistically
\ilya{I want a better word than ``statistically'' to indicate that
  it's doing something akin to Monte-Carlo integration (but I would
  like to reserve Monte-Carlo for MCMC).}.

\ilya{Something about how nested sampling distributes points according
  to the prior, not the posterior, and therefore can miss modes that
  are disfavored by prior.  Also, it does not exploit any information
  about the posterior to make the evidence calculation more
  efficient.}

In contrast to nested sampling, where computation of evidences is the
primary objective, Markov Chain Monte Carlo (MCMC) methods attempt to
stochastically produce parameter samples with density proportional to
the posterior probability distribution.  In MCMC techniques, the
primary target is an accurate estimate of the posterior distribution.
It is not straightforward to compute the model evidence from the MCMC
samples.  The most direct way to estimate the evidence for a model
from MCMC samples is to compute the harmonic-mean estimatator, but
this estimator of the evidence suffers from infinite
variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}; various
methods have been proposed to stabilize it \will{Citation}, but all
require some assumptions about the shape of the posterior distribution
that may not be satisfied in practice.

Reference \cite{Weinberg2009} gives a method for directly computing
the evidence integral using a kD-tree datastructure to decompose a
parameter space into boxes containing the MCMC sample points.  The
integral is approximated as a sum over box volumes.  This method is
promising, but it is not clear in general what statistical and
systematic errors it introduces and how these are affected by the
shape of the posterior distribution from which the MCMC samples.

@@

Implemented naively, RJMCMC has a significant drawback: because the
chain of samples must be Markovian, only the current sample is
available to the algorithm.  Each time an RJMCMC transitions between
models, the information about the good parameter values in the
previous model is lost; subsequent jumps into that model must ``start
fresh,'' and are correspondingly unlikely to be accepted, delaying
convergence of the RJMCMC sample chain.

Here we demonstrate a technique for improving the acceptance ratio of
inter-model jumps in an RJMCMC, leading to dramatically improved
convergence of RJMCMC sample chains.  The technique uses a kD-tree
datastructure to construct an approximation to each model's posterior
parameter distribution.  We propose jumps into each model drawn from
the approximation to the posterior.  Because jumps are proposed
preferentially to locations favored by the single-model posterior, the
RJMCMC compares ``good'' locations in parameter space across all the
models, and convergence is generally rapid.  The approximation to the
posterior distribution produced by the kD-tree is a constant-in-box
interpolation of the posterior, similar in spirit to the phase-space
density interpolants produced from N-body positions and momenta in
Refernce \cite{Ascasibar2005}.  The kD-tree interpolation is effective
in arbitrary numbers of dimension, and is quite space-efficient,
requiring $\order{N \log N}$ storage space and $\order{\log N}$ time
to produce each proposed jump, where $N$ is the number of single-model
MCMC samples used to construct the interpolation.  Reference
\cite{Littenberg2009} proposed an alternate method for producing
inter-model jumps in an RJMCMC that relies on a box decomposition of
parameter space, but uses fixed-sized boxes; because the box size is
fixed, the method cannot adapt to the local structure of the
posterior, and therefore becomes asymptotically less efficient than
the adaptive kD-tree decomposition in either acceptance ratio or
storage space as the parameter-space dimension increases or the
posterior becomes more peaked.  \ilya{Mention something (with
  reference) about the 10-way RJMCMC in the black-hole mass paper?}

The structure of this paper is as follows: in Section
\ref{sec:reversible-jump} we introduce in more detail the concept of a
Reversible Jump MCMC, and describe the fundamental difficulty with a
naive jump proposal in an RJMCMC.  In Section \ref{sec:kDTree} we
introduce the kD-tree datastructure used to decompose parameter space
into boxes for the interpolation.  In Section \ref{sec:efficiency} we
demonstrate the efficiency gains that are achieved from use of the
interpolated jump proposal.  In Section \ref{sec:examples} we give
examples of some other uses of the interpolated jump proposal in a
single-model MCMC context.  Finally, in Section \ref{sec:conclusion}
we offer a summary and some concluding remarks on the method.

@@

When selecting among alternative models, we are interested in finding
the model with the highest posterior probability $p(M_i|d)$.  However,
attempts to directly compute the evidence by performing the
integration in Eq.~\eqref{evidence} are generally very difficult in a
multi-dimensional, multi-modal parameter space when the likelihood has
to be evaluated numerically.  In particular, a grid-based integral
quickly becomes computationally unfeasible as the dimensionality of
$\vtheta$ exceeds a few.  The parameter space must typically be
explored in a stochastic manner before the evidence integral can be
computed.  There are several stochastic parameter-exploration
techniques focused directly on evidence computation (e.g., nested
sampling \cite{Skilling:2006} and MultiNest \cite{Feroz:2009}).  However, we
frequently want to compute the posterior PDFs within each model along
with the evidences for the various models.  One of the most common
techniques for computing posterior PDFs in the context of a model is
the Markov Chain Monte Carlo, which we now describe.

\subsection{MCMC} \label{sec:mcmc}

@@

One way to produce such a sequence of samples is via the
Metropolis-Hastings algorithm, first proposed in Reference
\cite{Metropolis:1953}, and later generalized by Hastings
\ilya{Citation?}:
\begin{enumerate}
\item Given a current state $\vtheta^{(j)}$, propose the next state
  $\vtheta^p$ by drawing from a jump proposal distribution with
  probability $Q(\vtheta^{(j)} \to \vtheta^p)$.
\item Compute the probability of accepting the proposed jump as
  \bel{eq:p-accept} 
p_{\textnormal{accept}} \equiv \min\Bigl(1,
\frac{p(\vtheta^p|d)}{p(\vtheta^{(j)}|d)} \frac{Q(\vtheta^p \to
  \vtheta^{(j)})}{Q(\vtheta^{(j)} \to \vtheta^p)} \Bigr).  
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<
  p_{\textnormal{accept}}$, accept the proposed jump, setting
  $\vtheta^{(j+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain
  at the same location in parameter space for the next step,
  $\vtheta^{(j+1)}=\vtheta^{(j)}$.
\end{enumerate}

>>>>

\IfFileExists{revtex4-1.cls}{\documentclass[prd, twocolumn, lengthcheck,
superscriptaddress, showpacs, letterpaper, nofootinbib]{revtex4-1}}{ 
\IfFileExists{revtex4.cls}{\documentclass[prd, preprint,
superscriptaddress, showpacs, letterpaper, nofootinbib]{revtex4}}{}
}

@@

\IfFileExists{apsrev4-1.bst}{\bibliographystyle{apsrev4-1}}
\IfFileExists{apsrev4.bst}{\bibliographystyle{apsrev4}}

@@

Selection among alternative theoretical models given an observed data
set is an important challenge in many areas of physics and astronomy.
In a Bayesian context, model selection involves computing the evidence
for each model given the data.  The model evidence is an integral of
the unnormalized posterior probability distribution over the model
parameter space, representing the probability of obtaining the data set within that model.
Models with larger evidence are preferred; the ratio of the evidences
of two models is the Bayes factor between them.  The product of the
Bayes factor and the ratio of prior probabilities for the two models yields
the odds ratio for the models.
%, the ratio of Bayesian probabilities for the two models given the observation data.

There are many ways to compute model evidences.  In low-dimensional
parameter spaces, the unnormalized posterior probability can be
evaluated on a grid or lattice and the integral can be performed
directly.  For many problems or models of interest, however, the
dimensionality of parameter space is too large to make this approach
practical, and stochastic sampling must be used.

%\ilya{Something about how nested sampling distributes points according to the prior, not the posterior, and therefore can miss modes that are disfavored by prior.  Also, it does not exploit any information about the posterior to make the evidence calculation more efficient.}

Markov Chain Monte Carlo (MCMC) methods attempt to
stochastically produce parameter samples with density proportional to
the posterior probability distribution.   In MCMC techniques, the primary target is an accurate estimate of the posterior distribution.  (We note that an alternative stochastic method for exploring a model parameter space, nested sampling \cite{Skilling:2004, Skilling:2006,Feroz:2009}, focuses on evidence computation rather than sampling the posterior probability density functions.)  \will{Will, are you happy with nested sampling in a parenthetical expression, without further comments?}
It is not straightforward to compute the model evidence from the MCMC
samples.  The most direct way to estimate the evidence for a model
from MCMC samples is to compute the harmonic-mean estimator, but
this estimator of the evidence suffers from infinite
variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  Various
methods have been proposed to stabilize the harmonic-mean estimator \will{Citation}, but all
require some assumptions about the shape of the posterior distribution
that may not be satisfied in practice.  MCMC implementations with parallel tempering \cite{EarlDeem:2005} allow for evidence computation via thermodynamic integration, but these can be computationally costly.

Reference \cite{Weinberg2009} gives a method for directly computing
the evidence integral from existing MCMC samples by using a kD-tree data structure to decompose a
parameter space into boxes containing the MCMC sample points.  The
integral is approximated as a sum over box volumes.  This method is
promising, but it is not clear in general what statistical and
systematic errors it introduces and how these are affected by the
shape of the posterior distribution from which the MCMC samples.

@@

Implemented naively, RJMCMC has a significant drawback: because the
chain of samples must be Markovian, only the current sample is
available to the algorithm.  Each time an RJMCMC transitions between
models, the information about the choices of parameter values in the
previous model is lost; subsequent jumps into that model must ``start
fresh,'' and are correspondingly unlikely to be accepted, delaying
convergence of the RJMCMC sample chain.

Here we introduce a technique for improving the acceptance ratio of
inter-model jumps in an RJMCMC, leading to dramatically improved
convergence of RJMCMC sample chains.  The technique uses a kD-tree
data structure to construct an approximation to each model's posterior
parameter distribution.  We draw jump proposals into the model from this approximation to its posterior.  Because jumps are proposed
preferentially to locations favored by the single-model posterior, the
RJMCMC compares ``good'' locations in parameter space across all the
models, and convergence is generally rapid.  We have successfully applied this RJMCMC technique to a 10-way model selection among alternative mass distribution models for black-hole X-ray binaries \cite{Farr2010}.

Reference \cite{Littenberg2009} proposed an alternate method for producing
inter-model jumps in an RJMCMC that relies on a box decomposition of
parameter space, but uses fixed-sized boxes.  The box size is
fixed, so the method cannot adapt to the local structure of the
posterior, and becomes asymptotically inefficient for high-dimensional parameter spaces or highly peaked posteriors.  Meanwhile, the approximation to the
posterior distribution produced by the kD-tree is a constant-in-box
interpolation of the posterior, similar in spirit to the phase-space
density interpolants produced from N-body positions and momenta in
Reference \cite{Ascasibar2005}.  The kD-tree interpolation is effective
in parameter spaces of arbitrary dimensionality, and is quite space-efficient,
requiring $\order{N \log N}$ storage space and $\order{\log N}$ time
to produce each proposed jump, where $N$ is the number of single-model
MCMC samples used to construct the interpolation.

The structure of this paper is as follows.  In Section
\ref{sec:reversible-jump} we introduce in more detail the concept of a
Reversible Jump MCMC, and describe the fundamental difficulty with a
naive jump proposal in an RJMCMC.  In Section \ref{sec:kDTree} we
introduce the kD-tree data structure used to decompose the parameter space
into boxes for interpolation.  In Section \ref{sec:efficiency} we
demonstrate the efficiency gains that are achieved from use of the
interpolated jump proposal.  In Section \ref{sec:examples} we give
examples of some other uses of the interpolated jump proposal, and suggest its utility in the context of a
single-model MCMC.  Finally, in Section \ref{sec:conclusion}
we offer a summary and some concluding remarks on the method.

@@

When selecting among alternative models, we are interested in finding
the model with the highest posterior probability $p(M_i|d)$.  However,
attempts to directly compute the evidence by performing the
integration in Eq.~\eqref{evidence} are generally very difficult in a
multi-dimensional, multi-modal parameter space when the likelihood has
to be evaluated numerically.  In particular, a grid-based integral
quickly becomes computationally unfeasible as the dimensionality of
$\vtheta$ exceeds a few.  The parameter space must typically be
explored in a stochastic manner before the evidence integral can be
computed.  There are several stochastic parameter-exploration
techniques focused directly on evidence computation (e.g., nested
sampling \cite{Skilling:2006} and MultiNest \cite{Feroz:2009}).  However, we
frequently want to compute the posterior PDFs within each model along
with the evidences for the various models.  One of the most common
techniques for computing posterior PDFs in the context of a model is
the Markov chain Monte Carlo, which we now describe.

\subsection{Markov chain Monte Carlo} \label{sec:mcmc}

@@

One way to produce such a sequence of samples is via the
Metropolis-Hastings algorithm, first proposed in Reference
\cite{Metropolis:1953}, and later generalized by Hastings \cite{Hastings:1970}:
\begin{enumerate}
\item Given a current state $\vtheta^{(j)}$, propose the next state
  $\vtheta^p$ by drawing from a jump proposal distribution with
  probability $Q(\vtheta^{(j)} \to \vtheta^p)$.
\item Compute the probability of accepting the proposed jump as
  \bel{eq:p-accept} 
p_{\textnormal{accept}} \equiv \min\Bigl(1,
\frac{p(\vtheta^p|d)}{p(\vtheta^{(j)}|d)} \frac{Q(\vtheta^p \to
  \vtheta^{(j)})}{Q(\vtheta^{(j)} \to \vtheta^p)} \Bigr).  
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<
  p_{\textnormal{accept}}$, accept the proposed jump, setting
  $\vtheta^{(j+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain
  at the same location in parameter space for the next step,
  $\vtheta^{(j+1)}=\vtheta^{(j)}$.
\end{enumerate}
